### 2020-03-01 doc/icfp20/conclusion.tex
wip

\section{Conclusion}
\label{sec:conclusion}

Text here.

>>>>

\section{Conclusion and Future Work}
\label{sec:conclusion}

We believe that functional programming, and in particular functional
programming with dependent types, is the most promising way to do software
engineering going forward. The languages and tools are perhaps not yet
as finely polished and rich as those typically used in production
software, but they are fundamentally far more powerful, and thus once
the infrastructure (and training of programmers) catches up should
become the languages of choice.

Music is an especially valuable domain in which to explore these
ideas. We have found it to be a microcosm of issues that arise
everywhere in software engineering: modularity, composability,
correctness, and data issues such as handling equivalent formulations of
the same data as well slightly different formulations. If we can
demonstrate that functional programming with dependent types works
well in the domain of music, the same techniques can be used in
software in general.

Data issues are particularly prevalent at API boundaries between
systems, which may internally represent the data in different formats
(satisfying modularity and abstraction) but then rely on tedious and
potentially error-prone conversions between the formats. In music
we see this in wanting to work sometimes with absolute pitch and others
with relative pitch (a pair of an octave and a pitch within that
octave), or choosing between
a pair of pitches and a \texttt{PitchInterval}. Also we would like to
reuse a function that just works on a pitch to work on a note, which
consists of a pitch plus a duration.

For now we have explicity converted between equivalent representations
and explictly lifted functions, to get a feel for the amount of work
required.  However in the world of dependent types there are known
research techniques for handling these cases, namely transporting
across equivalences as in Homotopy Type Theory~\citep{hottbook} and
ornaments~\citep{dagand-ornaments}. Cubical
Agda~\citep{vezzosi-cubical}, although still early in development,
should be an excellent tool for exploring the extent to which these
research ideas can be applied in a practical context.

Aside from these more fundamental ideas there is simply the work of
further extending the formalization of music theory to encompass
harmonic analysis, more advanced counterpoint, voice leading, and even
composition~\citep{schoenberg-fundamentals}. The results so far have
been encouraging, but much more needs to be done to find the right
abstractions to express the rules in the simplest, clearest and most
composeable way possible.
We expect the world of music theory to worthy of a long exploration.

### 2020-03-01 doc/icfp20/intro.tex

To describe rules of basic harmonic structure,
\citet{magalhaes-harmtrace} and its successors use dependent
types. (TODO: Give an example.) However Haskell currently has limited
support for dependent types, and requires many extentions and tricks
such as the use of singleton types~\citep{eisenberg-singleton} to
simulate a restricted form of dependent types. (TODO: fix wording)

@@

Full dependent types allow expression of predicate logic, and it is
tempting to take a standard textbook on music theory such as
\citet{piston-harmony} or \citet{aldwell2018harmony} and attempt to
formalize it in type theory. However this is difficult since most
``rules'' in music theory are not absolute, but rather suggestions of
varying degrees (themselves vaguely specified) of importance. Since
ultimately the resulting created music is more important than the
theory, an alternative is to attempt to indirectly deduce the rules
from music samples, using for example machine
learning~\citep{huang-cp}. However.... TODO
TODO: Perhaps create rules and weight them, then use ML.

Although formalization of a large amount of music theory is a goal,
for now we start with the modest task of expressing a small,
relatively strict rule set known as species
counterpoint~\cite{fux-cp}. TODO--more here
We also present preliminary work on harmonizing a melody using
a subset of rules based on \citet{piston-harmony}, contrasting with
existing work by \citet{koops-fharm}. Notably, since counterpoint and
harmony are not separate concepts but in fact deeply intertwined, we
are able to reuse counterpoint rules to help develop natural-sounding
harmonizations.

We make use of and extend~\citet{MusicTools}, a library of small tools
for analysis and synthesis of music written in Agda, with the goal
of eventually being a dependently-typed alternative to Euterpea. 
Previous work by ~\citet{cong-cp} using Music Tools expressed
first-species counterpoint with the Agda type system, so that an error
in counterpoint resulted in an Agda type error. Although this takes
advantage of Agda's native type checking, it has several
downsides. One is that the Agda errors may be difficult to interpret
as the corresponding musical error. Another is TODO: flexibility and
composability.

This is an experience report, and we highlight both the advantages and
disadvantages of using Agda. TODO: MORE

TODO: Talk about conclusion.

TODO: check that bib style is allowed.

>>>>

To describe rules of basic harmonic structure,
\citet{magalhaes-harmtrace} and its successors use dependent types,
for example to index chords by major or minor mode.  However Haskell
currently has limited support for dependent types, and requires many
extentions and tricks such as the use of singleton
types~\citep{eisenberg-singleton}.

@@

Full dependent types allow expression of predicate logic, and it is
tempting to take a standard textbook on music theory such as
\citet{piston-harmony} or \citet{aldwell2018harmony} and attempt to
formalize it in type theory. However this is difficult since most
``rules'' in music theory are not absolute, but rather suggestions of
varying degrees (themselves vaguely specified) of importance. Since
ultimately the resulting created music is more important than the
theory, an alternative is to attempt to indirectly deduce the rules
from music samples, using for example machine
learning~\citep{huang-cp}. However there is the danger that since
the rules were never explicit they could end up being violated with
poor results; for example parallel fifths and octaves that are
extremely rare (and almost always due to special allowed
circumstances) in Bach chorales used as training data resulted
in a much higher frequency of their use in generated
harmonizations~\citep{roberts-cp}.

Ultimately it seems some kind of mixture of rules and statistics may
be the best way to faithfully capture an effective theory of
music. One simple approach would be to develop a logic of rules and
their interactions, but attach weights to the rules which could be
determined via machine learning techniques.

Our long-term goal is formalization of a large amount of music theory.
As a first step we start with the modest task of expressing a small,
relatively strict rule set known as species
counterpoint~\cite{fux-cp}, intended for combining interdependent
melody lines to produce a pleasant-sounding result.

We make use of and extend~\citet{MusicTools}, a library of small tools
for analysis and synthesis of music written in Agda, with the goal of
eventually being a dependently-typed alternative to Euterpea.
Previous work by \citet{cong-cp} using Music Tools expressed
first-species counterpoint with the Agda type system, so that an error
in counterpoint resulted in an Agda type error. Although this takes
advantage of Agda's native type checking, it has several
downsides. One is that the Agda errors may be difficult to interpret
as the corresponding musical error. Another is that it is difficult to
keep the types both simple and flexible.

In this work we present an alternative representation of counterpoint
using what could be considered a custom musical type checker written
in Agda. This allows us to easily describe fine-grained rules, write
custom type errors, and then combine subsets of rules together for
larger-scale checking. Music can be easily combined with rules it
satisfies as a dependent record type. This record represents a proof
certificate that the music follows the asserted rules.
Rules can also easily be used in
other contexts, satisfying modularity and composability.

We also present preliminary work on harmonizing a melody using
a subset of rules based on \citet{piston-harmony}, contrasting with
existing work by \citet{koops-fharm}. Notably, since counterpoint and
harmony are not separate concepts but in fact deeply intertwined, we
are able to reuse counterpoint rules to help develop natural-sounding
harmonizations.

This is an experience report, and we highlight both the advantages and
disadvantages of using Agda for music theory. On one hand the
expressiveness of dependent types makes it easy and natural to
describe music theory rules. However we find the emphasis on proof
construction and particularly the extra work needed for decidable
equality can add an extra burden which is not always welcome. However
overall we feel the positives far outweigh the negatives, and in the
final section we describe future work which may help reduce the tedium.
